PineAPPL: combining EW and QCD corrections\\ for fast evaluation of LHC processes

### 2021-04-09 paper/sec-pineappl.tex
Correct mistake in sentence

\item \textsc{PineAPPL} has been explicitly interfaced to \textsc{mg5\_aMC} (v3+)~\cite{Alwall:2014hca,Frederix:2018nkq},\footnote{A version of \textsc{mg5\_aMC} interfaced to \textsc{PineAPPL} can be downloaded from \url{https://launchpad.net/mg5amc-pineappl/trunk}, and it will be included as standard in the next \textsc{mg5\_aMC} release.} see appendix~\ref{app:sample-runcard} for an example of a runcard.
The usage of \textsc{mg5\_aMC}(v3+) + \textsc{PineAPPL} is similar to that of \textsc{mg5\_aMC}(v2) + \textsc{aMCfast} + \textsc{APPLgrid}.
While the latter toolchain proved to be sufficiently efficient for NLO QCD calculations of typical kinematic distributions (with a few bins), this is no longer so when taking into account NLO EW calculations and, with more precise data, distributions that are typically more complex and have more bins.
Taking into account these two reasons, we find that the memory usage of \textsc{APPLgrid} is no longer satisfiable.
In particular, using \textsc{APPLgrid} for NLO QCD+EW\footnote{For this a modified version of \textsc{APPLgrid} was developed which added support for photon-initiated processes and EW corrections.} typically requires close to \SI{120}{\giga\byte} for high-mass Drell--Yan with 13 bins (see section~\ref{sec:dy-lepton-pair-production}).
The memory usage is substantially reduced to roughly \SI{1}{\giga\byte} after \enquote{optimisation} of the grids, i.e.\ an optimisation of the grid representation in memory.
However, this requires a two-step procedure: first, to produce an unoptimised grid in order to identify those parts where the cross section is either zero or extremely suppressed; second, after those parts are removed, to fill an optimised grid with a small number of bins.
\textsc{PineAPPL} avoids this by using a more space-efficient representation from the start. This leads to substantially faster run times, in particular for simple processes, because the grids do not need to be optimised 
and their combination is faster.

@@

In experiments, but also for many calculations where the phase-space integration is performed using MC techniques, finite statistics does not allow for the exact reconstruction of the dependence of the cross section on the observable $\mathcal{O}$.
Instead, it it sufficient to approximate the derivative using a piecewise-constant function,
\begin{equation}
W_{ab}^{(k,l,m,n)} \left( x_1, x_2, Q^2, \mathcal{O} \right) \approx \sum_{o=1}^M \frac{\Theta (\mathcal{O}_o^\mathrm{min} \le \mathcal{O} < \mathcal{O}_o^\mathrm{max})}{\mathcal{O}_o^\mathrm{max} - \mathcal{O}_o^\mathrm{min}} w_{ab}^{(k,l,m,n,o)} \left( x_1, x_2, Q^2 \right) \text{,}
\label{eq:differential-cross-section}
\end{equation}
which uses $M$ bins with limits $\{ \mathcal{O}_o^\mathrm{min}, \mathcal{O}_o^\mathrm{max} \}_{o=1}^M$ to partition a finite range of the observable,
\begin{equation}
\mathcal{O}_0^\mathrm{min} < \mathcal{O}_0^\mathrm{max} = \mathcal{O}_1^\mathrm{min} < \ldots < \mathcal{O}_{M-1}^\mathrm{max} = \mathcal{O}_M^\mathrm{min} < \mathcal{O}_M^\mathrm{max} \text{.}
\label{eq:bins-of-diff-xsection}
\end{equation}
The Ellis--Sexton scale $Q^2$, if chosen dynamically, depends on the phase space, however we assume the fractions $\xi_\mathrm{R}$ and $\xi_\mathrm{F}$ to be  phase-space constants in any case.
This allows for variations around the central scale choice $\xi_\mathrm{R} = \xi_\mathrm{F} = 1$, but it does not otherwise allow for arbitrary changes of the scale.
The terms with powers $m > 0$ and $n > 0$ vanish for the central scale choice and are only required for variations of the factorisation and renormalisation scales.
To estimate the perturbative QCD uncertainty --- no EW uncertainty is covered by this method --- one typically uses scale variations. Common prescriptions are
7-point and 9-points scale variations, which evaluate the cross section using
respectively the following values
  \begin{align}
    (\xi_\mathrm{R}, \xi_\mathrm{F})_{\mathrm{7-pt}}
    & \in
    \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr) \right\} \text{;}
    \label{eq:7pt}\\
    (\xi_\mathrm{R}, \xi_\mathrm{F})_{\mathrm{9-pt}}
    & \in
    \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr), \bigl( \tfrac{1}{2}, 2 \bigr), \bigl(2, \tfrac{1}{2} \bigr) \right\}  \text{.}
    \label{eq:9pt}
   \end{align}
The (asymmetric) uncertainties are then given as the minimum and maximum value (the envelope), measured from the central value $(1, 1)$.
As is clear from eq.~\eqref{eq:expansion}, the EW coupling $\alpha$ is assumed not to be a dynamically varying coupling, but instead a constant over phase space.
This, however, includes the most common choices of the coupling, which are (not necessarily in this order), $\alpha (0)$, $\alpha (M_\mathrm{Z})$, and $\alpha_{G_\mu}$.

>>>>

\item \textsc{PineAPPL} has been explicitly interfaced to \textsc{mg5\_aMC} (v3+)~\cite{Alwall:2014hca,Frederix:2018nkq},\footnote{A version of \textsc{mg5\_aMC} interfaced to \textsc{PineAPPL} can be downloaded from \url{https://launchpad.net/mg5amc-pineappl/trunk}, and it will be included as standard starting from the 3.0.4 release of \textsc{mg5\_aMC}.} see appendix~\ref{app:sample-runcard} for an example of a runcard.
The usage of \textsc{mg5\_aMC}(v3+) + \textsc{PineAPPL} is similar to that of \textsc{mg5\_aMC}(v2) + \textsc{aMCfast} + \textsc{APPLgrid}.
While the latter toolchain proved to be sufficiently efficient for NLO QCD calculations of typical kinematic distributions (with a few bins), this is no longer so when taking into account NLO EW calculations and, with more precise data, distributions that are typically more complex and have more bins.
Taking into account these two reasons, we find that the memory usage of \textsc{APPLgrid} is no longer satisfiable.
In particular, using \textsc{APPLgrid} for NLO QCD+EW\footnote{For this a modified version of \textsc{APPLgrid} was developed which added support for photon-initiated processes and EW corrections.} typically requires close to \SI{120}{\giga\byte} for high-mass Drell--Yan with 13 bins (see section~\ref{sec:dy-lepton-pair-production}).
The memory usage is substantially reduced to roughly \SI{1}{\giga\byte} after \enquote{optimisation} of the grids, i.e.\ an optimisation of the grid representation in memory.
However, this requires a two-step procedure: first, to produce an unoptimised grid in order to identify those parts where the cross section is either zero or extremely suppressed; second, after those parts are removed, to fill an optimised grid with a small number of bins.
\textsc{PineAPPL} avoids this by using a more space-efficient representation from the start. This leads to substantially faster run times, in particular for simple processes, because the grids do not need to be optimised 
and their combination is faster.

@@

In experiments, but also for many calculations where the phase-space integration is performed using MC techniques, finite statistics does not allow for the exact reconstruction of the dependence of the cross section on the observable $\mathcal{O}$.
Instead, it is sufficient to approximate the derivative using a piecewise-constant function,
\begin{equation}
W_{ab}^{(k,l,m,n)} \left( x_1, x_2, Q^2, \mathcal{O} \right) \approx \sum_{o=1}^M \frac{\Theta (\mathcal{O}_o^\mathrm{min} \le \mathcal{O} < \mathcal{O}_o^\mathrm{max})}{\mathcal{O}_o^\mathrm{max} - \mathcal{O}_o^\mathrm{min}} w_{ab}^{(k,l,m,n,o)} \left( x_1, x_2, Q^2 \right) \text{,}
\label{eq:differential-cross-section}
\end{equation}
which uses $M$ bins with limits $\{ \mathcal{O}_o^\mathrm{min}, \mathcal{O}_o^\mathrm{max} \}_{o=1}^M$ to partition a finite range of the observable,
\begin{equation}
\mathcal{O}_0^\mathrm{min} < \mathcal{O}_0^\mathrm{max} = \mathcal{O}_1^\mathrm{min} < \ldots < \mathcal{O}_{M-1}^\mathrm{max} = \mathcal{O}_M^\mathrm{min} < \mathcal{O}_M^\mathrm{max} \text{.}
\label{eq:bins-of-diff-xsection}
\end{equation}
The Ellis--Sexton scale $Q^2$, if chosen dynamically, depends on the phase space, however we assume the fractions $\xi_\mathrm{R}$ and $\xi_\mathrm{F}$ to be  phase-space constants in any case.
This allows for variations around the central scale choice $\xi_\mathrm{R} = \xi_\mathrm{F} = 1$, but it does not otherwise allow for arbitrary changes of the scale.
The terms with powers $m > 0$ and $n > 0$ vanish for the central scale choice and are only required for variations of the factorisation and renormalisation scales.
To estimate the perturbative QCD uncertainty --- no EW uncertainty is covered by this method --- one typically uses scale variations. Common prescriptions are
7-point and 9-points scale variations, which evaluate the cross section using
respectively the following values
  \begin{align}
    (\xi_\mathrm{R}, \xi_\mathrm{F})_{\mathrm{7-pt}}
    & \in
    \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr) \right\} \text{;}
    \label{eq:7pt}\\
    (\xi_\mathrm{R}, \xi_\mathrm{F})_{\mathrm{9-pt}}
    & \in
    \left\{ \bigl( 1, 1 \bigr), \bigl( \tfrac{1}{2}, \tfrac{1}{2} \bigr), \bigl( 2, 2 \bigr), \bigl( \tfrac{1}{2}, 1 \bigr), \bigl( 1, \tfrac{1}{2} \bigr), \bigl( 2, 1 \bigr), \bigl( 1, 2 \bigr), \bigl( \tfrac{1}{2}, 2 \bigr), \bigl(2, \tfrac{1}{2} \bigr) \right\}  \text{.}
    \label{eq:9pt}
   \end{align}
The (asymmetric) uncertainties are then given as the minimum and maximum value (the envelope), measured from the central value $(1, 1)$.
As is clear from eq.~\eqref{eq:expansion}, the EW coupling $\alpha$ is assumed not to be a dynamically varying coupling, but instead a constant over phase space.
This, however, includes the most common choices of the coupling, which are (not necessarily in this order), $\alpha (0)$, $\alpha (M_\mathrm{Z})$, and $\alpha_{G_\mu}$.
